Use of significance test logic by scientists in a novel reasoning task

### 2020-05-26 inst/manuscript/manuscript.Rmd
small changes to manuscript

[^1]: A formal statistical explanation showing that the task is difficult or impossible to perform using non-SST logic is given in Section 3 of Supplement A.

@@

@@

In addition to a random effect size, participants were randomly assigned to one of two transformations of the location/color test statistic from an underlying $z$ statistic. Of particular interest was how the transformation affected responding for the same visual deviation from the center. In one condition ("wide"), the null sampling distribution was visually about twice as wide as in the other ("narrow"). If participants were using the shuffle reports to interpret the data, as would be predicted if they were using SST logic, the transformation should affect their interpretation of the visual evidence: a visually-extreme point should be more discounted against the sampling distribution that is wider.

The effects of the manipulation are shown in figure \ref{fig:response-pvals}.

@@

Our findings echo other work showing that human reasoning can, under some conditions, be better than previously understood. [@Cosmides:Tooby:1992;@Gigerenzer:Hoffrage:1995]. Suggestions that SST be discontinued due to scientists' apparent misunderstandings may have been hasty. Of course, there may be other reasons why some favor abandoning SST, but our work shows that given the opportunity, scientists successfully deploy basic SST logic. In spite of scientists' real-life statistical behaviour often resembling a "ritual" [@Gigerenzer:etal:2004], when we eliminate the ritual --- no $p$ value, or any other familiar number, was offered --- they think statistically, very often arriving at the correct conclusion about the sign of the effect. Effective methodological reform in the sciences may be easier than abandoning significance testing, if we can harness scientists' already existing statistical competence.

@@

Participants were recruited via social media platforms such as Twitter and Facebook. All participants gave informed consent. Data inclusion criteria included sampling at least one shuffle report and experimental result, working in a scientific field, having at least some University education in science, and that it was their first time participating. Details are given in Supplement B.

@@

Each participant was randomly assigned to one of eight true effect sizes (from $\delta=0$ to $\delta=1$) and one of two evidence powers ("wide" $q=3$ or "narrow" $q=7$; see "Evidence Distributions" below).  The probability of being assigned $\delta=0$ was 25%, while the remaining effect sizes were equally probable at 11%. The probability of assignment to either evidence power was 50%.  Details are given in Table 1.1 in Supplement A.

After offering informed consent, participants read the cover story and instructions. During the instructions, the participant was introduced to the task through sampling random shuffle reports. After a brief recap of the instructions, participants performed the main task --- sampling either random shuffles or experiments --- until they made a decision about which, if either, elf group was faster. They were then asked several open-ended questions about their strategy, some informational questions (results in Supplement B) and debriefed.

@@

The evidence/horizontal ($x$) location test statistic presented to the participant was derived from a transformed $Z$ statistic:
\[
Z \sim \mbox{Normal}(\delta\sqrt{n/2}, 1)
\]
where $\delta$ is a true effect size (randomly assigned to each participant, from 0 to 1) and $n$ is the selected but unknown sample size (from 10 to 200 participants per group). $Z$ then transformed to the (-1,1) space:
\[
x = \mbox{sgn}(Z)\left[1 - \left(1 - F_{\chi_1^2}\left(Z^2\right)\right)^{\frac{1}{q}}\right],\ -1\leq x \leq 1.
\]
where $F_{\chi_1^2}$ is the cumulative distribution function of a $\chi^2_1$ random variable, and $q\in\{3,7\}$ was randomly assigned for each participant. $x=-1$ represented the left edge of the interface, $x=0$ the middle, and $x=1$ the right edge. The setting of $q$ determined how spread out the test statistic was on the display. This arbitrary transformation was done to ensure that the test statistic's distribution was unfamiliar to the participant. See Supplement A for more details, including graphical depictions of the evidence distributions.

@@

We determined the coding scheme and independently categorized the first 20 participant, discussing the source of disagreements. After categorizing the remaining participants, some disagreements were resolved through mutual agreement, and a discussion between the authors was had over what caused the disagreements. The remainder of the disagreements were re-coded separately, and a final round of discussion resolved the remaining disagreements. Participants' codings described in detail in Supplement B.

>>>>

[^1]: A formal statistical explanation showing that the task is difficult or impossible to perform using non-SST logic is given in Section 3 of [Supplement A](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html).

@@

Our experiment was contructed such that the only way to assess the evidence in the data was by comparison to a null sampling distribution: either the one provided by the random shuffle reports, or a simpler null that assumes that the evidence will favor one team or the other with 50% probability. Thus, the information afforded only the information in a $p$ value, but it was not described as such, and participants had to *discover* for themselves how to use the informaiton.

@@

In addition to a random effect size, participants were randomly assigned to one of two transformations of the location/color test statistic from an underlying $z$ statistic. Of particular interest was how the transformation affected responding for the same visual deviation from the center. In one condition ("wide"), the null sampling distribution was visually about twice as wide as in the other ("narrow").

The effects of the manipulation are shown in Figure \ref{fig:response-pvals}. The two experimental conditions used different arbitrary monotone mappings from the underlying $Z$-statistic to the visual space. Intuitively, this would be like deciding to use $Z^3$ instead of $Z$ in all $Z$ tests; one would need to adjust the significance criteria to account for the cubing (e.g., use $|1.96^3|=7.53$ instead of $|1.96|$ for a $\alpha=0.05$ level test), but the underlying test remains the same. The manipulation changes only the visual impression of the sampling distributions, allowing us to see how sensitive their responses are to the null sampling distribution as represented by the random shuffle reports.

If participants were using the shuffle reports to interpret the data, as would be predicted if they were using SST logic, the transformation should affect their interpretation of the visual evidence: a visually-extreme point should be more discounted against the sampling distribution that is wider. When we break down responses by the *visual* extremeness of the evidence, responses in two conditions should appear different; when we break down responses by  *statistical* extremeness (i.e., the $p$ value) responses in the two conditions should appear very similar, because the visual manipulation is irrelevant, given the $p$ value.

@@

There are some major differences between our methods and the methods of previous demonstrations that scientists fail when reasoning about SST. We have already mentioned the fact that previous demonstrations typically use survey vignettes, as opposed to our more engaged experimental method. A second difference is our sample which is a substantially larger and more diverse convenience sample than previous demonstrations. It is possible that the recent so-called "replication crisis" [@Baker:2016] has raised awareness among scientists of deficits of statistical reasoning. If this increase in awareness accounts for our more optimistic findings, this would be good news for statistical educators.

Our findings echo other work showing that human reasoning can, under some conditions, be better than previously understood. [@Cosmides:Tooby:1992;@Gigerenzer:Hoffrage:1995]. Suggestions that SST be discontinued due to scientists' apparent misunderstandings may have been hasty. Of course, there may be other reasons why some favor abandoning SST, but our work shows that given the opportunity, scientists successfully deploy basic SST logic. In spite of scientists' real-life statistical behaviour often resembling a "ritual" [@Gigerenzer:etal:2004], when we eliminate the ritual --- no $p$ value, or any other familiar number, was offered --- they think statistically, very often arriving at the correct conclusion about the sign of the effect. Effective methodological reform in the sciences may be easier than abandoning significance testing, if we can harness scientists' already-existing statistical competence.

@@

Participants were recruited via social media platforms such as Twitter and Facebook. All participants gave informed consent. Data inclusion criteria included sampling at least one shuffle report and experimental result, working in a scientific field, having at least some University education in science, and that it was their first time participating. Details are given in [Supplement B](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html).

@@

Each participant was randomly assigned to one of eight true effect sizes (from $\delta=0$ to $\delta=1$) and one of two evidence powers ("wide" $q=3$ or "narrow" $q=7$; see "Evidence Distributions" below).  The probability of being assigned $\delta=0$ was 25%, while the remaining effect sizes were equally probable at 11%. The probability of assignment to either evidence power was 50%.  Details are given in Table 1.1 in [Supplement A](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html).

After offering informed consent, participants read the cover story and instructions. During the instructions, the participant was introduced to the task through sampling random shuffle reports. After a brief recap of the instructions, participants performed the main task --- sampling either random shuffles or experiments --- until they made a decision about which, if either, elf group was faster. They were then asked several open-ended questions about their strategy, some informational questions (results in [Supplement B](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html)) and debriefed.

@@

The evidence/horizontal ($x$) location test statistic presented to the participant was derived from a transformed $Z$ statistic:
\[
Z \sim \mbox{Normal}(\delta\sqrt{n/2}, 1)
\]
where $\delta$ is a true effect size (randomly assigned to each participant, from 0 to 1) and $n$ is the selected but unknown sample size (from 10 to 200 participants per group). $Z$ then transformed to the (-1,1) space:
\[
x = \mbox{sgn}(Z)\left[1 - \left(1 - F_{\chi_1^2}\left(Z^2\right)\right)^{\frac{1}{q}}\right],\ -1\leq x \leq 1.
\]
where $F_{\chi_1^2}$ is the cumulative distribution function of a $\chi^2_1$ random variable, and $q\in\{3,7\}$ was randomly assigned for each participant. $x=-1$ represented the left edge of the interface, $x=0$ the middle, and $x=1$ the right edge. The setting of $q$ determined how spread out the test statistic was on the display. This arbitrary transformation was done to ensure that the test statistic's distribution was unfamiliar to the participant. See [Supplement A](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html) for more details, including graphical depictions of the evidence distributions.

@@

We determined the coding scheme and independently categorized the first 20 participant, discussing the source of disagreements. After categorizing the remaining participants, some disagreements were resolved through mutual agreement, and a discussion between the authors was had over what caused the disagreements. The remainder of the disagreements were re-coded separately, and a final round of discussion resolved the remaining disagreements. The coding of participants' responses is described in detail in [Supplement B](https://richarddmorey.github.io/Morey_Hoekstra_StatCognition/articles/man_supp.html).
